Capturing the geometric structure of our experiences and how we remember them

### 2018-07-23 boneyard.tex
checking in after brain figure

>>>>

To test the observation that participants' recall trajectories generally matched the shape of the video trajectory, we fit participants' recall data to a model comprised of $n$ line segments that were connected by a subset of the average recall event coordinates. We fixed the endpoints of the model at the average first and last recall event, but fit the placement/length of rest of the line segments to the data. Using a cross-validated model fitting procedure (see ~\ref{sec:methods} for details), we searched for the optimal number of line segments. The best fitting model (12 line segments) can be seen in Fig.~\ref{fig:linesegs}. This model represents the set of consecutive line segments that best describe all participants recall event models. Notably, the optimal model comprises a subset of the total recall events (~roughly 1/3). % include a measure of the variance described by the model?

We hypothesized that the points connecting the line segments (referred to as ``hinge-points'') represent critical scenes in the storyline. To visualize the semantic content of each hinge-point, we created wordles from the average topic vectors for each video event (left side), where the size of the word is proportional to its ``activation'' in the video event relative to other video events (Fig.~\ref{fig:linesegs}, see ~\ref{sec:methods} for analysis details). We performed the same analysis for the average recall topic vectors (right side). Notably, the text in the wordles is distinct across the hinge-points, but but similar between video and recall within hinge-point.

\subsubsection{Line segment analysis}
This analysis tested whether recall dynamics across participants could be explained by a smaller number of consecutive line segments.  To test this idea, we ran an analysis similar to a piecewise linear regression, with a few additional constraints. First, the endpoints of the line segments were constrained to the average recall event coordinates. Second, we did not consider solutions with neighboring events (i.e. there must be at least 1 event between line segments).  This greatly reduced the search space, making the analysis computationally tractable.  Third, we considered models with different numbers of line segments (2-15).  To fit the model, we randomly split participants into 2 groups (i.e. training and test). For the training group, we (exhaustively) searched for $n$ consecutive line segments (from 2-15) that best explained the recall data. The cost function was given by the grand average euclidean distance of each participant's recall events to the closest line segment in the model. Then, the test group was also fit to this model.  We repeated this procedure 100 times, randomly splitting the participants on each iteration.  We chose $n$ (i.e. the number of line segments) by selecting the value with the lowest test error.  Finally, we refit the model including all subjects to give the line segment model that best explained the data.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_projs.pdf}
\caption{\small \textbf{Individual participant recall event model embeddings.} Each participant's recall event model embedded into a 2D space using the UMAP algorithm. The arrows on the lines connecting the recall events represents the direction of the sequence of recalls.  The color of the dots represent the most likely (i.e. most correlated) video event. The dark black line represents the best fitting line segment model (same for each participant).}
\label{fig:ind_embeddings}
\end{figure}

### 2018-07-23 main.tex
checking in after brain figure

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/5_linesegs.pdf}
\caption{\small \textbf{Recall shape analysis and video/recall content.} A 2-dimensional embedding of the best fitting line segment model.  The points outlined in black represent and connected by black line segments represent the critical recall events (or ``hinge-points'') of the best fitting model colored according to their matching video event.  The points not outlined in black are individual participant recall event topic vectors colored by the most similar video event. The kernel density plot represents the distribution across all points.  The word clouds for hinge-point (left: video; right: recall) were created by plotting the top 100 words from topic model that were ``activated'' during a particular event.}
\label{fig:linesegs}
\end{figure}

To test the observation that participants' recall trajectories generally matched the shape of the video trajectory, we fit participants' recall data to a model comprised of $n$ line segments that were connected by a subset of the average recall event coordinates. We fixed the endpoints of the model at the average first and last recall event, but fit the placement/length of rest of the line segments to the data. Using a cross-validated model fitting procedure (see ~\ref{sec:methods} for details), we searched for the optimal number of line segments. The best fitting model (12 line segments) can be seen in Fig.~\ref{fig:linesegs}. This model represents the set of consecutive line segments that best describe all participants recall event models. Notably, the optimal model comprises a subset of the total recall events (~roughly 1/3). % include a measure of the variance described by the model?

We hypothesized that the points connecting the line segments (referred to as ``hinge-points'') represent critical scenes in the storyline. To visualize the semantic content of each hinge-point, we created wordles from the average topic vectors for each video event (left side), where the size of the word is proportional to its ``activation'' in the video event relative to other video events (Fig.~\ref{fig:linesegs}, see ~\ref{sec:methods} for analysis details). We performed the same analysis for the average recall topic vectors (right side). Notably, the text in the wordles is distinct across the hinge-points, but but similar between video and recall within hinge-point.

@@

\subsubsection{Line segment analysis}
This analysis tested whether recall dynamics across participants could be explained by a smaller number of consecutive line segments.  To test this idea, we ran an analysis similar to a piecewise linear regression, with a few additional constraints. First, the endpoints of the line segments were constrained to the average recall event coordinates. Second, we did not consider solutions with neighboring events (i.e. there must be at least 1 event between line segments).  This greatly reduced the search space, making the analysis computationally tractable.  Third, we considered models with different numbers of line segments (2-15).  To fit the model, we randomly split participants into 2 groups (i.e. training and test). For the training group, we (exhaustively) searched for $n$ consecutive line segments (from 2-15) that best explained the recall data. The cost function was given by the grand average euclidean distance of each participant's recall events to the closest line segment in the model. Then, the test group was also fit to this model.  We repeated this procedure 100 times, randomly splitting the participants on each iteration.  We chose $n$ (i.e. the number of line segments) by selecting the value with the lowest test error.  Finally, we refit the model including all subjects to give the line segment model that best explained the data.

\subsubsection{Topic vector word clouds}
 We created word clouds to visualize the themes contained in the ``hinge-point'' recall events of the model. One component of the topic model comprises a words (???) by topics (100) matrix (R), where the rows represent the weighting of a given word on each topic. Each event (video and recall) was associated with a 1 by 100 topic vector (v) representing the mixture of topics contained in the event. To find words associated with a given event, we computed the dot product between R and v, which gave a 1 by \# of words vector where the values represent the `activation' of each of in the event. We normalized these vectors by subtracting the average `activation' across all events. We performed this analysis for each of the line segment model `hinge-points' (i.e. the average recall events connecting the line segments). We also performed this analysis for the video events at the hinge-point locations. Finally, we created word clouds by extracting the top 100 words and plotting them where the size of the word is proportional to its `activation' in the event.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_projs.pdf}
\caption{\small \textbf{Individual participant recall event model embeddings.} Each participant's recall event model embedded into a 2D space using the UMAP algorithm. The arrows on the lines connecting the recall events represents the direction of the sequence of recalls.  The color of the dots represent the most likely (i.e. most correlated) video event. The dark black line represents the best fitting line segment model (same for each participant).}
\label{fig:ind_embeddings}
\end{figure}

>>>>

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Recall shape analysis and video/recall content.} A 2-dimensional embedding of the best fitting line segment model.  The points outlined in black represent and connected by black line segments represent the critical recall events (or ``hinge-points'') of the best fitting model colored according to their matching video event.  The points not outlined in black are individual participant recall event topic vectors colored by the most similar video event. The kernel density plot represents the distribution across all points.  The word clouds for hinge-point (left: video; right: recall) were created by plotting the top 100 words from topic model that were ``activated'' during a particular event.}
\label{fig:topics}
\end{figure}

\begin{figure}[]
\centering
\includegraphics[width=1\textwidth]{figs/6_brains.pdf}
\caption{\small \textbf{Recall model searchlight analysis.} A 2-dimensional embedding of the best fitting line segment model.  The points outlined in black represent and connected by black line segments represent the critical recall events (or ``hinge-points'') of the best fitting model colored according to their matching video event.  The points not outlined in black are individual participant recall event topic vectors colored by the most similar video event. The kernel density plot represents the distribution across all points.  The word clouds for hinge-point (left: video; right: recall) were created by plotting the top 100 words from topic model that were ``activated'' during a particular event.}
\label{fig:topics}
\end{figure}

@@

\subsubsection{Topic vector word clouds}
 We created word clouds to visualize the themes contained in the recall events. One component of the topic model comprises a words (XX) by topics (100) matrix (R), where the rows represent the weight of a given word in each topic.  To find words that were maximally associated with a particular event vector, we computed the dot product between R and v, which gave a 1 by \# of words vector where the values represent the `activation' of each of in the event. Then, we created word clouds by extracting the top $n$ words and plotting them where the size of the word is proportional to its `activation' in the event.

In the first analysis (Fig.~\ref{fig:topics}a/b), we quantified the most and least remembered topics/words throughout the entire video by computing a weighted average over all recall events, where the weights were proportional to memory for each recall event. To measure memory for each event, for each participant we computed the correlation between the video event vector and the closest recall event vector. We then averaged these correlation values across participants. We then computed a weighted average of all video events using the correlation values as weights. Next, we computed the dot product between this weighted-average video event vector and the R matrix (described in the paragraph above) to get ``activations'' for each word. Finally, we plotted the top 200 words where the size of the word is proportional to its ``activation''. To get the least remembered topics/words, we performed the same analysis but inverted memory weights.

In the second analysis (Fig.~\ref{fig:topics}c), we created wordles for the top/bottom 3 remembered video events indexed by the average correlation values (Fig.~\ref{fig:topics}a).  To get the ``activations'' for words associated with the video events, we computed the dot product between the video event vector and the R matrix. The same procedure was used to get word activations for the recall events. We then plotted the top 200 words for the top/bottom 3 recalled events.

 \subsection{fMRI analyses}
 Seventeen participants watched the first 50 minutes of Episode 1 of BBC's Sherlock. The video was split into two parts of approximately equal length (946 and 1030 TRs). All data were preprocessed and transformed to 3mm MNI space as described in the paper. Data were zscored across time at every voxel. 6mm smoothing was applied.
 Files are cropped so that all video-viewing data are aligned across participants, and all recall data are aligned to the scene timestamps below. The cropping includes a constant 3-TR (4.5 sec) shift to correct for hemodynamic lag.

 \subsubsection{Searchlight analysis}
 Our multivariate analyses were designed to capture brain regions whose timepoint-by-timepoint correlational structure mirrors the correlational structure of the participant-specific recall topic models during video viewing. To correct for non-linearities between the viewing time and recall time, for each participant we used DTW to temporally align the matrices. The algorithm recovers a path of coordinates that would bring the video and recall model in maximal temporal alignment. We used this path to warp the fMRI data and the recall model into temporal alignment (separately for each participant). Then, we conducted a searchlight analysis (5x5x5 voxel cube) where for each cube, we correlated the model timepoint-by-timepoint correlation matrix with the neural correlation matrix. To aggregate across participants, we Fisher's z-transformed the correlations and then averaged.  To assess significance, we recomputed this group analysis 100 times, but randomly phase shifted the model (by \# of timepoints - 1) by the same amount for each participant but different amounts for each permutation to build a null distribution of correlation values. Finally, we thresholded the group averaged correlation maps where the `real' correlation value for a given voxel exceeded the 95th percentile of the null distribution.

@@
